---
title: "6122_hw1"
author: "Gian Zlupko"
date: "2023-01-30"
output: html_document
---
###  HW 1 

Assignment: 
* Solve with R the five exercises from Chapter 1.8, pages 23-24:
* 1.1, 1.2, 1.3, 1.4, 1.5


Libaries Used
```{r, warning = FALSE}
library(tidyverse) 
```



Load data set for Table 1.1
```{r}
data_1 <-
  structure(list(individual = 1:10, sex = structure(c(2L, 2L, 2L,
    2L, 2L, 1L, 1L, 1L, 1L, 1L), .Label = c("Female", "Male"), class = "factor"),
    age = c(21L, 43L, 22L, 86L, 60L, 16L, NA, 43L, 22L, 80L),
    IQ = c(120L, NA, 135L, 150L, 92L, 130L, 150L, NA, 84L, 70L
    ), depression = structure(c(2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L,
    1L, 1L), .Label = c("No", "Yes"), class = "factor"), health = structure(c(3L,
    3L, 1L, 4L, 2L, 2L, 3L, 1L, 1L, 2L), .Label = c("Average",
    "Good", "Very good", "Very poor"), class = "factor"), weight = c(150L,
    160L, 135L, 140L, 110L, 110L, 120L, 120L, 105L, 100L)), .Names = c("individual",
    "sex", "age", "IQ", "depression", "health", "weight"), class = "data.frame", row.names = c(NA, -10L))

# view first 5 rows 
head(data_1) 

# check the data types in the data
str(data_1) 
```

#### 1.1: *Find the correlation matrix and covariance matrix of the data in Table 1.1*

Note: I subset the data to remove the categorical variables like sex
```{r}
health_subset <- data_1 %>% 
  select(-c(sex, depression, health))   

# covariance matrix
cov(health_subset) 

```

Correlation matrix 
```{r}
cor(health_subset) 
```


#### 1.2: *Fill in the missing values in Table 1.1 with appropriate mean values, and recalculate the correlation matrix of the data*

To estimate the means

```{r}
# subset the df to identify instances where value = NA
# then, use the mean() function to fill in the missing value 

# age: 
health_subset$age[is.na(health_subset$age)]<-mean(health_subset$age,na.rm=TRUE)
# IQ 
health_subset$IQ[is.na(health_subset$IQ)]<-mean(health_subset$IQ,na.rm=TRUE)

# eeight 
health_subset$weight[is.na(health_subset$weight)]<-mean(health_subset$weight,na.rm=TRUE)
```

Now I can recalculate the correlation matrix for the data 

```{r}
cor(health_subset) 

```


#### *1.3: Examine both the normal probability plots of each variable in the archaeology data in Table 1.3 and the chi-square plot of the data. Do the plots suggest anything unusual about the data?* 

```{r}
# load the archaeology data from Table 1.3 from the HSAUR2 package 
data("pottery", package = "HSAUR2")
head(pottery) 
```


Normal probability plots 
```{r}
# first I remove kiln, which is stored as a categorical variable (factor type) in R
pottery_1 = pottery %>% select(-kiln)
layout(matrix(1:9, nc = 3))
sapply(colnames(pottery_1), function(x){ 
  qqnorm(pottery_1[[x]], main = x)
  qqline(pottery_1[[x]])
  })
```

[insert comment]

Chi-square plot of the data 
```{r}

```




























