---
title: "6122_hw1"
author: "Gian Zlupko"
date: "2023-01-30"
output: html_document
---
###  HW 1 

Assignment: 
* Solve with R the five exercises from Chapter 1.8, pages 23-24:
* 1.1, 1.2, 1.3, 1.4, 1.5


Libaries Used
```{r, warning = FALSE, output = FALSE}
library(tidyverse) 
```



Load data set for Table 1.1
```{r}
data_1 <-
  structure(list(individual = 1:10, sex = structure(c(2L, 2L, 2L,
    2L, 2L, 1L, 1L, 1L, 1L, 1L), .Label = c("Female", "Male"), class = "factor"),
    age = c(21L, 43L, 22L, 86L, 60L, 16L, NA, 43L, 22L, 80L),
    IQ = c(120L, NA, 135L, 150L, 92L, 130L, 150L, NA, 84L, 70L
    ), depression = structure(c(2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L,
    1L, 1L), .Label = c("No", "Yes"), class = "factor"), health = structure(c(3L,
    3L, 1L, 4L, 2L, 2L, 3L, 1L, 1L, 2L), .Label = c("Average",
    "Good", "Very good", "Very poor"), class = "factor"), weight = c(150L,
    160L, 135L, 140L, 110L, 110L, 120L, 120L, 105L, 100L)), .Names = c("individual",
    "sex", "age", "IQ", "depression", "health", "weight"), class = "data.frame", row.names = c(NA, -10L))

# view first 5 rows 
head(data_1) 

# check the data types in the data
str(data_1) 
```
 
n = 10 
q = 7 


#### 1.1: *Find the correlation matrix and covariance matrix of the data in Table 1.1*

Note: I subset the data to remove the categorical variables like sex
```{r}
health_subset <- data_1 %>% 
  select(-c(sex, depression, health))   

# covariance matrix
cov(health_subset) 

```

Correlation matrix 
```{r}
cor(health_subset) 
```


#### 1.2: *Fill in the missing values in Table 1.1 with appropriate mean values, and recalculate the correlation matrix of the data*

To estimate the means

```{r}
# subset the df to identify instances where value = NA
# then, use the mean() function to fill in the missing value 

# age: 
health_subset$age[is.na(health_subset$age)]<-mean(health_subset$age,na.rm=TRUE)
# IQ 
health_subset$IQ[is.na(health_subset$IQ)]<-mean(health_subset$IQ,na.rm=TRUE)

# eeight 
health_subset$weight[is.na(health_subset$weight)]<-mean(health_subset$weight,na.rm=TRUE)
```

Now I can recalculate the correlation matrix for the data 

```{r}
cor_res <- cor(health_subset) 
# round the correlations to two decimal places 
round(cor_res, 2)
```


#### *1.3: Examine both the normal probability plots of each variable in the archaeology data in Table 1.3 and the chi-square plot of the data. Do the plots suggest anything unusual about the data?* 

```{r}
# load the archaeology data from Table 1.3 from the HSAUR2 package 
data("pottery", package = "HSAUR2")
head(pottery) 
```
Note: Kiln is nominal.

Normal probability plots 
```{r}
# first I remove kiln, which is stored as a categorical variable (factor type) in R
pottery_1 = pottery %>% select(-kiln)
layout(matrix(1:9, nc = 3))
sapply(colnames(pottery_1), function(x){ 
  qqnorm(pottery_1[[x]], main = x)
  qqline(pottery_1[[x]])
  })
```

Visual inspection of the chi-square plots suggest that the majority of the variables deviate from linearity. For example, in the plots above, BaO and Na2O appear to be the only two variables in the data set that follow normal distributions. In addition, many of the variables show evidence of outliers. Outliers are especially pronounced in the variables TiO2 and Na2O. 

Next, I inspect 

Chi-square plot of the data 
```{r}
x <- pottery_1
cm <- colMeans(x)
S <- cov(x)

d <- apply(x, 1, function(x) t(x - cm) %*% solve(S) %*% (x - cm))


plot(qc <- qchisq((1:nrow(x) - 1/2) / nrow(x), df = 7), 
     sd <- sort(d),
     xlab = expression(paste(chi[7]^2, " Quantile")), 
     ylab = "Ordered distances", xlim = range(qc) * c(1, 1.1))
oups <- which(rank(abs(qc - sd), ties = "random") > nrow(x) - 3)
text(qc[oups], sd[oups] - 1.5, names(oups))
abline(a = 0, b = 1)

```
The chi-square plot shows that the ordered distances do not fall on the straight line. These results may suggest that the observations do not arise from a multivariate normal distribution. In addition, there appear to be multiple outliers that have greater distance values from the mean vector. 




```{r}
###################################################
### Exercise 1.4
###################################################
s <- c(3.8778,
       2.8110,  2.1210,
       3.1480,  2.2669,  2.6550,
       3.5062,  2.5690,  2.8341,   3.2352)
S <- diag(4)
S[!lower.tri(S)] <- s
S <- S + t(S)
diag(S) <- diag(S) / 2
```





```{r}
###################################################
### Exercise 1.5
###################################################
X <- matrix(
 c(3, 4, 4, 6, 1,
   5, 1, 1, 7, 3,
   6, 2, 0, 2, 6,
   1, 1, 1, 0, 3,
   4, 7, 3, 6, 2,
   2, 2, 5, 1, 0,
   0, 4, 1, 1, 1,
   0, 6, 4, 3, 5,
   7, 6, 5, 1, 4,
   2, 1, 4, 3, 1), ncol = 5)

```


















