---
title: "6122_hw5"
author: "Gian Zlupko"
date: "2023-03-23"
output: html_document
---


Ex. 5.1 Show how the result Σ = ΛΛ⊤ + Ψ arises from the assumptions of uncorrelated factors, independence of the specific variates, and indepen- dence of common factors and specific variances. What form does Σ take if the factors are allowed to be correlated?

```{r}

```



Ex. 5.2 Show that the communalities in a factor analysis model are unaffected by the transformation Λ∗ = ΛM.

```{r}


```



Ex. 5.3 Give a formula for the proportion of variance explained by the jth factor estimated by the principal factor approach.

```{r}

```

Ex. 5.4 Apply the factor analysis model separately to the life expectancies of men and women and compare the results.

```{r}

```


Ex. 5.5 The correlation matrix given below arises from the scores of 220 boys in six school subjects: (1) French, (2) English, (3) History, (4) Arithmetic, (5) Algebra, and (6) Geometry. Find the two-factor solution from a max- imum likelihood factor analysis. By plotting the derived loadings, find an orthogonal rotation that allows easier interpretation of the results.

```{r}

```




Ex. 5.6 The matrix below shows the correlations between ratings on nine statements about pain made by 123 people suffering from extreme pain. Each statement was scored on a scale from 1 to 6, ranging from agreement to disagreement. 


(a) Perform a principal components analysis on these data, and exam- ine the associated scree plot to decide on the appropriate number of components.
(b) Apply maximum likelihood factor analysis, and use the test described in the chapter to select the necessary number of common factors.
```{r}

```



























