---
title: "6122_hw5"
author: "Gian Zlupko"
date: "2023-03-23"
output: html_document
---


Ex. 5.1 Show how the result Σ = ΛΛ⊤ + Ψ arises from the assumptions of uncorrelated factors, independence of the specific variates, and independence of common factors and specific variances. What form does Σ take if the factors are allowed to be correlated?


Answer: 

The result Σ = ΛΛ⊤ + Ψ arises from the assumptions of uncorrelated factors, independence of the specific variates, and independence of common factors and specific variances, which can be represented mathematically as:

Σ = ΛΛ⊤ + Ψ

Where Λ is the matrix of common factors, Λ⊤ is the transpose of Λ, and Ψ is the matrix of independent specific variances.

If the factors are allowed to be correlated, the form of Σ is given by:

Σ = ΛΛ⊤ + Ψ + Φ

Where Φ is the matrix of correlations between the factors.



```{r}
# Load mtcars data
data(mtcars)

# Compute matrix of common factors Λ
Lambda <- cor(mtcars)

# Compute matrix of specific variances Ψ
Psi <- diag(apply(mtcars, 2, var))

# Compute matrix of correlations Φ
Phi <- cor(mtcars) - Lambda

# Compute matrix of full covariance Σ
Sigma <- Lambda %*% t(Lambda) + Psi + Phi

Sigma
```



*Ex. 5.2 Show that the communalities in a factor analysis model are unaffected by the transformation Λ∗ = ΛM.*




*Ex. 5.3 Give a formula for the proportion of variance explained by the jth factor estimated by the principal factor approach.*


Answer: Proportion of Variance Explained by jth Factor = (Eigenvalue of jth Factor/Sum of all Eigenvalues) * 100


*Ex. 5.4 Apply the factor analysis model separately to the life expectancies of men and women and compare the results.*


First create the data set that appears in the chapter. 

```{r}
"life" <- structure(.Data = list(c(63., 34., 38., 59., 56., 62., 50., 65., 56., 69., 65., 64., 56., 60., 61., 49., 59., 63., 59., 65., 65., 64.,
64., 67., 61., 68., 67., 65., 59., 58., 57.)
, c(51., 29., 30., 42., 38., 44., 39., 44., 46., 47., 48., 50., 44., 44., 45., 40., 42., 44., 44., 48., 48., 63., 43., 45., 40., 46., 45., 46., 43., 44., 46.)
, c(30., 13., 17., 20., 18., 24., 20., 22., 24., 24., 26., 28., 25., 22., 22., 22., 22., 23., 24., 28., 26., 21.,21., 23., 21., 23., 23., 24., 23., 24., 28.)
, c(13., 5., 7., 6., 7., 7., 7., 7., 11., 8., 9., 11., 10., 6., 8., 9., 6., 8., 8., 14., 9., 7., 6., 8., 10., 8., 8., 9., 10., 9., 9.) , c(67., 38., 38., 64., 62., 69., 55., 72., 63., 75., 68., 66., 61., 65., 65., 51., 61., 67., 63., 68., 67., 68., 68., 74., 67., 75., 74., 71., 66., 62., 60.), c(54., 32., 34., 46., 46., 50., 43., 50., 54., 53., 50., 51., 48., 45., 49., 41., 43., 48., 46., 51., 49., 47., 47., 51., 46., 52., 51., 51., 49., 47., 49.) , c(34., 17., 20., 25., 25., 28., 23., 27., 33., 29., 27., 29., 27., 25., 27., 23., 22., 26., 25., 29., 27., 25., 24., 28., 25., 29., 28., 28., 27., 25., 28.) , c(15., 6., 7., 8., 10., 14., 8., 9., 19., 10., 10., 11., 12., 9., 10., 8., 7., 9., 8., 13., 10., 9., 8., 10., 11., 10., 10., 10., 12., 10., 11.)), class = "data.frame", names = c("m0", "m25", "m50", "m75", "w0", "w25", "w50", "w75"), row.names = c("Algeria", "Cameroon", "Madagascar", "Mauritius", "Reunion", "Seychelles", "South Africa (C)", "South Africa (W)", "Tunisia", "Canada", "Costa Rica", "Dominican Rep.", "El Salvador", "Greenland", "Grenada", "Guatemala","Honduras", "Jamaica", "Mexico", "Nicaragua", "Panama", "Trinidad (62)", "Trinidad (67)","United States (66)", "United States (NW66)", "United States (W66)", "United States (67)", "Argentina", "Chile", "Colombia", "Ecuador"))

# inspect data set
head(life) 

# store male and female subsets 
m_life <- life[, 1:4]
f_life <- life[, 5:8]
```


To apply the EFA model on the male data set, I fit a model with k = 1 as specifying any larger values of k led to the following error ("Error in factanal(m_life, factors = 2) :
2 factors are too many for 4 variable")


```{r}
factanal(m_life, factors = 1)
```

Next, I do the same on the female data set. 

```{r}
factanal(f_life, factors = 1) 
```

Results show that a single latent factor accounted for more of the overall variance in the female data set than a single latent factor did in the male data set. For both outputs, we see that all four variables load strongly onto the latent factor. 




Ex. 5.5 The correlation matrix given below arises from the scores of 220 boys in six school subjects: (1) French, (2) English, (3) History, (4) Arithmetic, (5) Algebra, and (6) Geometry. Find the two-factor solution from a maximum likelihood factor analysis. By plotting the derived loadings, find an orthogonal rotation that allows easier interpretation of the results.

First I manually create the correlation matrix in order to load it into R. 

```{r}
french <- c(1.00, .44, .41, .29, .33, .25)
english <- c(0.44, 1.00, .35, .35, .32, .33)
history <- c(.41, .35, 1.00, .16, .19, .18)
arithmetic <- c(.29, .35, .16, 1.00, .59, .47)
algebra <- c(.33, .32, .19, .59, 1.00, .46)
geometry <- c(.25, .33, .18, .47, .46, 1.00)

cor_mat <- rbind(french, english, history, arithmetic, 
                 algebra, geometry)

colnames(cor_mat) <- c("french", "english", "history", 
                       "arithmetic", "algebra", "geometry")

cor_mat

```




Next, I create a two-factor model with maximum likelihood and orthogonal rotation .


```{r}
factanal(covmat = cor_mat, factors = 2, method = "mle")

fa_subjects <- factanal(covmat = cor_mat, factors = 2, method = "mle") 
fa_subjects 

```


Plot the factor loadings 

```{r}
loadings <- fa_subjects$loadings
loadings[, 1]

plot(fa_subjects$loadings[,1], fa_subjects$loadings[,2], xlab = "Factor 1", ylab = "Factor 2")
text(fa_subjects$loadings[,1], fa_subjects$loadings[,2], abbreviate(rownames(cor_mat), 5), cex = cex)
```


Now I will re-run the EFA model but instead of the varimax rotation I use promax rotatation. 

```{r}
factanal(covmat = cor_mat, factors = 2, rotation = "promax", method = "mle")
```


The results indicate that, with promax rotation, the individual subjects load more cleanly onto the two latent factors. Specifically, the math subjects dominate the first factor and the humanities subjects dominate the second factor. Moreover, with the promax rotation, the humanities negatively load onto the first factor indicating that they are negatively correlated with this factors and, instead, are more related to the second factor. 


E*x. 5.6 The matrix below shows the correlations between ratings on nine statements about pain made by 123 people suffering from extreme pain. Each statement was scored on a scale from 1 to 6, ranging from agreement to disagreement.* 


(a) Perform a principal components analysis on these data, and exam- ine the associated scree plot to decide on the appropriate number of components.
(b) Apply maximum likelihood factor analysis, and use the test described in the chapter to select the necessary number of common factors.
```{r}

```
















Ch. 5 code 
```{r}
library(MVA)
demo("Ch-EFA")
```










