---
title: "hw9"
author: "Gian Zlupko"
date: "2023-04-26"
output: html_document
---

### HUDM 6122 HW9 
#### Gian Zlupko 

Exercises 7.1, 7.2 and 7.3 on pages 223-224.




*#### 7.1 The matrix below shows the correlations between ratings on nine statements about pain made by 123 people suffering from extreme pain. Each statement was scored on a scale from 1 to 6, ranging from agreement to disagreement.*

*Fit a correlated two-factor model in which questions 1,3,4, and 8 are assumed to be indicators of the latent variable Doctor’s Responsibility and questions 2,5,6, and 7 are assumed to be indicators of the latent variable Patient’s Responsibility. Find a 95% confidence interval for the correlation between the two latent variables.*


Load in data manually 

```{r}
cor_mat = matrix(c(1.00, -.04, .61, .45, .03, -.29, -.30, .45, .30, 
         -.04, 1.00, -.07, -.12, .49, .43, .30, -.31, -.17, 
         .61, -.07, 1.00, .59, .03, -.13, -.24, .59, .32, 
         .45, -.12, .59, 1.00, -.08, -.21, -.19, .63, .37, 
         .03, .49, .03, -.08, 1.00, .47, .41, -.14, -.24, 
         -.29, .43, -.13, -.21, .47, 1.00, .63, -.13, -.15, 
         -.3, .3, -.24, -.19, .41, .63, 1.00, -.26, -.29, 
         .45, -.31, .59, .63, -.14, -.13, -.26, 1.00, .40, 
         .30, -.17, .32, .37, -.24, -.15, -.29, .40, 1.00), nrow = 9, ncol = 9)

cor_mat

rownames(cor_mat) <- c("item1", "item2", 
                            "item3", "item4",
                            "item5", "item6", "item7", "item8", "item9")
colnames(cor_mat) <- c("item1", "item2", 
                            "item3", "item4",
                            "item5", "item6", "item7", "item8", "item9")
cor_mat
```


Fit the CFA model as specified by the instructions in the problem

```{r, message = F}
library(lavaan) 

# model specification 
mod1 <- ' 
doc_responsibility =~ item1 + item3 + item4 + item8 
pat_responsibility =~ item2 + item5 + item6 + item7 ' 

# fit CFA model 
mod1_fit <- cfa(mod1, sample.cov = cor_mat, sample.nobs = 123) 

# inspect standardized item loadings
```

The correlation between the two factors is r = -.305, which we can see above in the model summary output, under covariances, in the Std.lv column.

Next, while the question does not ask us to, we can inspect the model fit with function below: 
```{r}
fitMeasures(mod1_fit)
```

Results revealed the following fit statistics: 

- Chi-Square: 63.75
- Chi-square p-value: 0.00
- CFI: 0.87
- TLI: 0.81
- RMSEA: 0.138 [0.10, .17]
- SRMR: .077

First, the results show that chi-square is significantly large, and the associated p-value is small, leading us to reject the null that sigma and the sample covariance matrix are equivalent. Basically, this is an indication that the conceptual model does not adequately represent the relationships in the data. Next, we see that the fit statistics are sub-optimal. Specifically, CFI, RMSEA and SRMR are all below their recommended cut-off values (Hu & Bentler, 1999; Kline, 2009). Thus, each of these outputs indicates that the model fits OK but not that well. 


Finally, for fun, we can plot the path diagram below using the `semPlots` package and can ask for the standardized parameter estimates. 
```{r, message = F}
library(semPlot) 

semPaths(object = mod1_fit, what = "paths", whatLabels = "std") 
```


*#### Ex. 7.2 For the stability of alienation example, fit the model in which the measurement errors for anomia in 1967 and anomia in 1971 are allowed to be correlated.*

First, I load the data, which is provided in the chapter demo:

Note the data that is provided in the chapter demo does not have all of the variables that are shown in the SEM pictured on pg. 220 of the book (yet again an issue with the book!) We can simply run a SEM with the variables given and still provide correlated errors (correlated residuals) for the anomia 1967 and anomia 1971 variables. 

```{r}
alienation <- matrix(c(11.839, 6.947, 6.819, 4.783, -3.834, -21.899, 6.947, 9.364, 5.091, 5.028, -3.889, -18.831, 6.819, 5.091,12.532, 7.495, -3.841, -21.748, 4.783, 5.028, 7.495, 9.986, -3.625, -18.775, -3.834,-3.889,-3.841,-3.625,  9.60,   35.522, -21.899,-18.831,-21.748,-18.775,35.522,450.283), ncol = 6, byrow = TRUE)

rownames(alienation) <- colnames(alienation) <- c("Anomia67", "Powles67", "Anomia71", "Powles71", "Educ", "SEI")

# inspect the data 
alienation
```


Next, I specify the SEM model below. 

```{r}

alien_mod <- '  
# measument portion of SEM
Alienation67 =~ Anomia67  + Powles67
Alienation71 =~ Powles71 + Anomia71

# paths 

Alienation71 ~ Alienation67 

# correlated residual
Anomia67 ~~ Anomia71

' 

# fit the SEM model on the covariance data
sem1_fit <- lavaan::sem(model = alien_mod, 
                        sample.cov = alienation, sample.nobs = 932)

# inspect model fit statistics and output 
summary(sem1_fit, standardized = T)

```



Plot the path diagram of the model

```{r}
semPaths(object = sem1_fit, 
         whatLabels = "std")
```

The path diagram provided by the `semPaths` library is not very easy to read in this case, but in the image, we can see the correlated residuals for the



*#### Ex. 7.3 Meyer and Bendig (1961) administered the five Thurstone Primary Mental Ability tests, verbal meaning (V), space (S), reasoning (R), numerical (N), and word fluency (W), to 49 boys and 61 girls in grade 8 and again three and a half years later in grade 11. The observed correlation matrix is shown below. Fit a single-factor model to the correlations that allows the factor at time one to be correlated with the factor at time two.*


Load the data 

```{r}

thurstone <- " 
1
0.37 1
0.42  0.33 1
0.53  0.14 0.38  1
0.38  0.10  0.20  0.24 1
0.81  0.34  0.49  0.58 0.32  1
0.35  0.65 0.20 -0.04 0.11  0.34  1
0.42  0.32 0.75  0.46  0.26 0.46 0.18 1
0.40  0.14 0.39  0.73  0.19 0.55 0.06 0.54 1
0.24  0.15 0.17  0.15  0.43 0.24 0.15 0.20  0.16 1
"


thurstone_cov_mat <- getCov(thurstone, 
                            names = c("V1", "S1", "R1", "N1", "W1", 
                                      "V2", "S2", "R2", "N2", "W2"))

thurstone_cov_mat
```




The specification for a single factor model (also known as a unidimensional model) is as follows: 
```{r}

single_factor <- '
factor1 =~ V1 + S1 + R1 + N1 + W1 + V2 + S2 + R2 + N2 + W2

# correlated residuals
V1 ~~ V2
S1 ~~ S2
R1 ~~ R2
N1 ~~ N2
W1 ~~ W2

'

# fit the single factor model
thurstone_cfa_fit <- cfa(model = single_factor,
                         sample.cov = thurstone_cov_mat, 
                         sample.nobs = 110)

# inspect the model output
summary(thurstone_cfa_fit, standardized = TRUE, fit.measure = TRUE) 

```

We can see from the fit statistics that the model fits fairly well according to conventional cut-off recommendations (Hu & Bentler, 1999; Kline, 2009). RMSEA is nearing the high side by the 95% CI for RMSEA falls within the recommended range. 

- CFI: .95
- TLI: .92
- RMSEA: .08
- SRMR: .06

Chi square is significant, suggestion that the model does not exactly reproduce the covariances in the data, but overall the model fits well. 

```{r}
semPaths(thurstone_cfa_fit, what = "std") 
```


We can see from the path diagram that the items at Time 1 (or 8th grade) were allowed to correlate with the items at Time 2 (11th grade). This likely helped with model fit. For example, we can fit a similar model below but without the correlated residuals that the textbook asked for and compare the fit of each model. 

Model with no correlated residuals: 
```{r}
single_factor_no_corr <- '
factor1 =~ V1 + S1 + R1 + N1 + W1 + V2 + S2 + R2 + N2 + W2

# correlated residuals - none
'

# fit the single factor model
thursone2 <- cfa(model = single_factor_no_corr,
                         sample.cov = thurstone_cov_mat, 
                         sample.nobs = 110)

# inspect the model output
summary(thursone2, standardized = TRUE, fit.measure = TRUE) 
```


As expected, we can see that the model without correlated residuals fit much worse than the first model, which relaxed this contraint. 


Finally, while the chapter asks for 1 factor model, we can also explore fitting a two factor model, one for time 1 and time 2, and we can compare this model's fit to the model that the authors requested. 

Two factor model: 

```{r}
two_factor <- '
factor1 =~ V1 + S1 + R1 + N1 + W1 
factor2 =~ V2 + S2 + R2 + N2 + W2

# correlated residuals
V1 ~~ V2
S1 ~~ S2
R1 ~~ R2
N1 ~~ N2
W1 ~~ W2

'

# fit the single factor model
two_factor_fit <- cfa(model = two_factor,
                         sample.cov = thurstone_cov_mat, 
                         sample.nobs = 110)

# inspect the model output
summary(two_factor_fit, standardized = TRUE, fit.measure = TRUE) 
```

```{r}
semPaths(two_factor_fit, what = "std") 
```



This model arguably fits better than the first, single factor model. We can test whether one model fits better than the other with a chi-square difference test:

```{r}
lavTestLRT(thurstone_cfa_fit, two_factor_fit) 
```

The test is marginally significant but technically the difference in chi-square between the models is not significant at the .05 level, nor is the actual difference in chi-square very large (3.61). Thus, the researcher should choose the model that best suits their theoretical understanding of the constructs. 



