---
title: "midterm_study_code"
author: "Gian Zlupko"
date: "2023-03-07"
output: html_document
---


Code from Ch. 1 - Multivariate Data Analysis


```{r}
library(MVA) 
demo("Ch-MVA")
```





#### Pearson Correlation and Covariance

Note: use = complete.obs argument will discard the entire row if an NA is present.

```{r}

# correlation matrix, rounded to 2 decimals 
data_cor <- round(cor(mtcars, use = "complete"), 2)
data_cor
```


```{r}
# cov matrix rounded to 2 decimals
data_cov <- round(cov(mtcars, use = "complete"), 2) 

```



Convert a covariance matrix into a correlation matrix 


```{r}
mat_one <- matrix(c(3.8778, 2.8110, 3.1480, 3.5062,
                     2.8110, 2.1210, 2.2669, 2.5690,
                     3.1480, 2.2669, 2.6550, 2.8341,
                     3.5062, 2.5690, 2.8341, 3.2352),4,4, byrow=T)


# convert the matrix, 'mat_one', into a correlation matrix 
# use round function
round(cov2cor(mat_one), 2)

```






#### for loop 

Simple for loop function in R. Below, the loop writes over the values in the first three columns with the respective column mean. 

```{r}


for (cols in c(1,2,3)) { 
  
  mtcars[, cols] <- mean(mtcars[, cols])
  
}

# inspect the first three columns where the values were overwritten 
head(mtcars) 
```



#### Normal Probability Plots 


```{r}
library(HSAUR2) 

# create single normal probability plot 
qqnorm(pottery[, 1], main = "AL203") 

```


Create a grid of normal probability plots in data set 
```{r}

# create normal probability plots for each variable 

sapply(colnames(pottery)[1:9], function(x) { 
  
  qqnorm(pottery[[x]], main = x)
  qqnorm(pottery[[x]])
  
  })
```



#### Create a distance matrix

Euclidean
```{r}
mat_one

# use dist() function
dist(mat_one, method = "euclidean") 
```


Manhattan

```{r}
dist(mat_one, method = "manhattan")
```





