---
title: "mds_lab"
author: "Gian Zlupko"
date: "2023-03-01"
output: html_document
---








in principle, we ask R to make as many dimensions as possible because we want to choose how many dimensions to retain. The k argument in cmdscale() allows us to set the max number of dimensions. We can choose n-1 to choose the max number of dimensions. 


```{r}
mtcars

# create distance matrix 
d = dist(mtcars) 


# now apply MDS to distance matrix 
# where k is n-1 dimensions that are used to represent the dimensions 
cmdscale(d, k = 9, eig = TRUE) 

```



We can also store the MDS model to be able to view its outputs
```{r}

mds1 <- cmdscale(d, k = 9, eig = TRUE)

mds1$eig

# calculate the cumulative sum of the eigenvalues 

cumsum(abs(mds1$eig)) / sum(abs(mds1$eig)) 
```

Once we have inspected the eigenvalues, we can re-fit MDS with a new value of k for the number of dimensions that we want that was informed by the eigenvalues. 





Now, we can build a dist matrix using manhattan distance 
```{r}

# specify that we use manhattan distance in the dist() function
mds_2 <- cmdscale(dist(mtcars, method = "manhattan"),k = nrow(mtcars) - 1, eig = TRUE)
```












