---
title: "multiple correspondence analysis (MCA)"
author: "Gian Zlupko"
date: "2023-04-19"
output: html_document
---

```{r, message = F}
library(tidyverse)
```




Multiple correspondence analysis (MCA) is used when the variables are categorical. It is like a PCA for categorical data. 


Load data sets from local directory
```{r}

# manually select txt file from other directory 
food <- read.delim(file.choose())
cars <- read.delim("cars.txt", row.names = NULL)
bachelors <- read.delim(file.choose())
crime <- read.delim(file.choose()) 

# store csv in directory for later use
#write.csv(food, "food_data.csv")
#write.csv(bachelors, "bachelor.csv")
#write.csv(crime, "crime.csv")
#write.csv(cars, "cars.csv")

# load data sets 

cars <- read.csv('cars.csv') 
crime <- read.csv('crime.csv')
bachelor <- read.csv('bachelor.csv')
food <- read.csv('food_data.csv')

```


Cars data set 
```{r}
head(cars) 
# for loop to convert all columns to factor type 
for (i in 1:ncol(cars)) { 
  
  cars[, i] = as.factor(cars[, i])
  
}

```

Apply MCA
```{r}
mca_cars <- MCA(cars) 
mca_cars$eig
plot(mca_cars) 

categories_cars <- apply(cars_cat, 2, function(x) nlevels(as.factor(x)))
```


```{r}
# data frames for ggplot
mca_cars_vars_df = data.frame(mca_cars$var$coord, Variable = rep(names(categories_cars), 
    categories_cars))

mca_cars_obs_df = data.frame(mca_cars$ind$coord)

# plot of variable categories
ggplot(data = mca_cars_vars_df, aes(x = Dim.1, y = Dim.2, label = rownames(mca1_vars_df))) + 
    geom_hline(yintercept = 0, colour = "gray70") + geom_vline(xintercept = 0, 
    colour = "gray70") + geom_text(aes(colour = Variable)) + ggtitle("MCA plot of variables using R package FactoMineR")
```





#### Homoegeneity Analysis

Using the `Homals` package. 

Similar to CA but provides optimal scaling for different types of variables so that you can do a dim reduction with data that are measured on different scales (e.g. categorial, continuous, etc) 

```{r}
library(homals) 
require(homals, quietly = T)
mca5 = homals(newtea, ndim = 5, level = "nominal")

mca5
```

```{r}
tea
```






#### MCA Tutorial 

The following MCA tutorial follows the demonstration provided in the following link on RPubs: 
https://rpubs.com/gaston/MCA


Load FactoMineR package and tea data set from the package.
```{r}
library(FactoMineR) 
data(tea) 
head(tea) 
dim(tea) 
```


```{r}
# select these columns
newtea = tea[, c("Tea", "How", "how", "sugar", "where", "always")]

cats = apply(newtea, 2, function(x) nlevels(as.factor(x)))
# see how many categories we have within each variable 
cats
```


Apply MCA 
```{r}
# apply MCA
mca1 = MCA(newtea, graph = FALSE)

# table of eigenvalues
mca1$eig
```


The largest eigenvalue is 1 when dealing with categorical data because, 

when we look at the loadings corresponding to the first dimension, they will all be either positive or negative 1. 



```{r}
# column coordinates
head(mca1$var$coord)
```



```{r}
# data frames for ggplot
mca1_vars_df = data.frame(mca1$var$coord, Variable = rep(names(cats), 
    cats))
mca1_obs_df = data.frame(mca1$ind$coord)

# plot of variable categories
ggplot(data = mca1_vars_df, aes(x = Dim.1, y = Dim.2, label = rownames(mca1_vars_df))) + 
    geom_hline(yintercept = 0, colour = "gray70") + geom_vline(xintercept = 0, 
    colour = "gray70") + geom_text(aes(colour = Variable)) + ggtitle("MCA plot of variables using R package FactoMineR")
```

We can also superimpose a graphic display of both the observations and the categories. Moreover, since some individuals will be overlapped, we can use density curves to see those zones that are highly concentrated.


```{r}
# MCA plot of observations and categories
ggplot(data = mca1_obs_df, aes(x = Dim.1, y = Dim.2)) + geom_hline(yintercept = 0, 
    colour = "gray70") + geom_vline(xintercept = 0, colour = "gray70") + geom_point(colour = "gray50", 
    alpha = 0.7) + geom_density2d(colour = "gray80") + geom_text(data = mca1_vars_df, 
    aes(x = Dim.1, y = Dim.2, label = rownames(mca1_vars_df), colour = Variable)) + 
    ggtitle("MCA plot of variables using R package FactoMineR") + scale_colour_discrete(name = "Variable")
```
Interpretation: looking at the colors, we can see that some of the variables have high values on either of the dimensions/factors. The green 'other' variable is high on Dim2 and the variable 'tea shop' is very high on Dim1. 




```{r}
# default biplot in FactoMineR
plot(mca1)
```








Variable Clustering 

```{r}
library(ClustOfVar) 

str(food) 
food_num = food %>% select(-c(Workertype, X, ID)) 

food_var_cluster <- hclustvar(food_num) 
plot(food_var_cluster) 

```






